---
Name: Madison Rollins
title: 'Geog6300: Lab 6'
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Regression

```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(tidyverse)
```

**Overview:**
This lab focuses on regression techniques. You'll be analyzing the association of various physical and climatological characteristics in Australia with observations of several animals recorded on the citizen science app iNaturalist.

###Data and research questions###

Let's import the dataset.

```{r}
lab6_data<-st_read("data/aus_climate_inat.gpkg")
```

The dataset for this lab is a 1 decimal degree hexagon grid that has aggregate statistics for a number of variables:

* ndvi: NDVI/vegetation index values from Landsat data (via Google Earth Engine). These values range from -1 to 1, with higher values indicating more vegetation.
* maxtemp_00/20_med: Median maximum temperature (C) in 2000 or 2020 (data from SILO/Queensland government)
* mintemp_00/20_med: Median minimum temperature (C) in 2020 or 2020 (data from SILO/Queensland government)
* rain_00/20_sum: Total rainfall (mm) in 2000 or 2020 (data from SILO/Queensland government)
* pop_00/20: Total population in 2000 or 2020 (data from NASA's Gridded Population of the World)
* water_00/20_pct: Percentage of land covered by water at some point during the year in 2000 or 2020
* elev_med: Median elevation (meters) (data from the Shuttle Radar Topography Mission/NASA)

There are also observation counts from iNaturalist for several distinctively Australian animal species: the central bearded dragon, the common emu, the red kangaroo, the agile wallaby, the laughing kookaburra, the wombat, the koala, and the platypus.

Our primary research question is how the climatological/physical variables in our dataset are predictive of the NDVI value. We will build models for 2020 as well as the change from 2000 to 2020. The second is referred to as a "first difference" model and can sometimes be more useful for identifying causal mechanisms.

###Part 1: Analysis of 2020 data###

We will start by looking at data for 2020. 

**Question 1** _Create histograms for NDVI, max temp., min temp., rain, and population, and water in 2020 as well as elevation. Based on these graphs, assess the normality of these variables._

```{r}
variables_2020 <- c("ndvi_20_med", "maxtemp_20_med", "mintemp_20_med", "rain_20_sum", "pop_20", "water_20_pct", "elev_med")

for (var in variables_2020) {
  hist(lab6_data[[var]], main = paste(var), xlab = var, col = "red", breaks = 30)}
```

{Most of the variables in this analysis deviate from a normal distribution trend. Elevation and temperature have shown distributions closer to normality but still have slight skewing. Variables such as rainfall, population, and water percentage showed the highest level of skew with extreme deviations from normality.}

**Question 2** _Use tmap to map these same variables using Jenks natural breaks as the classification method. For an extra challenge, use `tmap_arrange` to plot all maps in a single figure._

```{r}
library(tmap)
tm_list <- lapply(variables_2020, function(var) {
  tm_shape(lab6_data) +
    tm_polygons(var, style = "jenks", palette = "viridis", title = var) +
    tm_layout(
      title = paste(var),
      title.position = c("center", "top"),  
      legend.outside = TRUE,               
      legend.outside.position = "left",
      legend.frame = TRUE)})

tmap_arrange(plotlist = tm_list)
```


**Question 3** _Based on the maps from question 3, summarise major patterns you see in the spatial distribution of these data from any of your variables of interest. How do they appear to be associated with the NDVI variable?_

{Variables such as rainfall, water percentage, and moderate temperatures are positively associated with higher NDVI values, meaning their are favorable conditions for vegetation growth. Areas with high maximum temperatures and low rainfall located in the central and western regions show a lower NDVI scale, meaning the condtions are less favorable for vegetation growth in these areas. Population density does not show a very clear relationship with the NDVI, naturally, more urban areas tend to have less vegetation which doesnt doesnt have to be based on the surrounding environment.}

**Question 4** _Create univariate models for each of the variables listed in question 1, with NDVI in 2020 as the dependent variable. Print a summary of each model. Write a summary of those results that indicates the direction, magnitude, and significance for each model coefficient._

```{r}
uni_var_2020 <- lapply(variables_2020, function(var) {
  model <- lm(ndvi_20_med ~ lab6_data[[var]], data = lab6_data)
  summary(model)})
uni_var_2020
```

{Most variables are highly significant (NDVI, max temp, min temp, rain, population, and elevation), with p-values less than 0.001, except for variable 6 (Water in 2020), which is not statistically significant due to having the p-value 0.249. Rain, population, and elevation show that the independent variable had a very small impact on the NDVI. The highest influence on the NDVI was found in the rain variable with a r squared of 0.4273 or (42.73%), while the lowest is in the water in 2020 variable with a r squared of 0.001862 or (0.186%). This data shows that the variables with higher r squared percentages fit the regression model better than the variables with a lower r squared percentage. So, the higher r squared values (observed in the NDVI, max temp, min temp, rain, population) means there is less variation between observed data in the fitted values where as a lower r squared value (observed in the water 2020 and elevation) means there was bigger differences or more variation between data and the fit. 

**Question 5** _Create a multivariate regression model with the variables of interest, choosing EITHER max or min temperature (but not both) You may also choose to leave out any variables that were insignificant in Q4. Use the univariate models as your guide. Call the results._

```{r}
multi_var_2020 <- lm(ndvi_20_med ~ maxtemp_20_med + rain_20_sum + pop_20 + water_20_pct + elev_med, data = lab6_data)
summary(multi_var_2020)
```

**Question 6** _Summarize the results of the multivariate model. What are the direction, magnitude, and significance of each coefficient? How did it change from the univariate models you created in Q4 (if at all)? What do the R2 and F-statistic values tell you about overall model fit?_

{The model helps explain a good chunk of the changes happening within the NDVI. The high R-value (0.6397, adjusted 0.6372) and the F-statistic (252.1 with 5 and 710 DF) suggest that the selected variables are suitable for modeling an NDVI.

Intercept: The given estimate is (0.4591) which means that if all the independent variables are at 0, the predicted NDVI is 0.4591. This result is highly significant due to having a p-value of (2x10^-16).

maxtemp_20_med: This variable has a negative relationship.The maximum temperature shows a 1 degree Celsius increase which can be associated with a the -0.01173 decrease found within the NDVI. The p-value shows high significance (p < 2x10^-16).

Rain_20_sum: The variable has a positive relationship with a value of (8.464x10^-7) which means that rainfall slightly increases within the NDVI. The  result is highly significant with a p-value of (2x10^-16).

pop_20: This variable has a positive relationship with a value of (2.873x10^-7) which means that a one-person increase in population creates an increase in the NDVI. The result is slightly significant with a p-value of (p = 0.00613).

water_20_pct: This variable has a negative relationship with a value of (-0.03387) which means that an increase in water coverage creates a decrease in NDVI. This result is not significant with a p-value of (0.7275).

elev_med:  This variable has a positive relationship with a value of (0.0001215) which means that an increase in elevation creates an increase in NDVI. This result is highly significant with a p-valye of (1.28x10^-10).}

**Question 7** _Use a histogram and a map to assess the normality of residuals and any spatial autocorrelation. Summarise any notable patterns that you see._

```{r}
lab6_data$residuals <- residuals(multi_var_2020)

ggplot(lab6_data, aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line()

shapiro.test(lab6_data$residuals)

ggplot(lab6_data, aes(x = residuals)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Residuals", x = "Residuals", y = "Frequency") +
  theme_minimal()

tm_shape(lab6_data) +
  tm_dots("residuals", size = 0.2)
```

{Most of the residuals are around zero or very low, meaning they are a good fit. There are slight deviations in normality, with a longer left tail and a few outliers on both sides. The presence of outliers suggests the model might have difficulty handling extreme cases but the residual clustering outcome indacates an overall good performance for the majority of observations.The outliers on the map are represented by the red and green dots because they reflect the ranges furthest from zero. These outliers are mostly present in coastal areas and could be affected by human or environmental factors.}

**Question 8** _Assess any issues with multicollinearity or heteroskedastity in this model using the techniques shown in class. Run the appropriate tests and explain what their results show you._

```{r}
library(car)
vif(multi_var_2020)

ncvTest(multi_var_2020)
```

{The values range from 1.08 to 1.17, which are all below the typical threshold for potential multicollinearity analysis. This means multicollinearity is not ideal to use for this model, as the varaibles are not strongly correlated. The NCV test has a chi-square value of 287.39, with 1 DF and a p-value less than (2.22 × 10^-16). This indicates a highly significant results. The test also shows that the assumption of homoscedasticity is not met, which means the spread of the residuals isn’t the same across the predicted values. This suggests the presence of heteroscedasticity in the model.}

**Question 9** _How would you summarise the results of this model in a sentence or two? In addition, looking at the full model and your diagnostics, do you feel this is a model that provides meaningful results? Explain your answer._

{This model provides meaningful results that can help predict NDVI trends and the variance of individual variables. I found the results of the maps to be the most meaningful as a geographer because they incorproate the spatial connection between the data and the natural enviornment. The outliers on the map highlighted areas of concern or possible data inconsistencies. These inconsistences (specifically around the coast) could provide insight into possible causes such as humans or certain environmental factors. The heteroskedasticity emphasizes possible limitations with the standard errors and significance tests. This is more a question of their accuracy. Adjustments can be made such as using standard errors or alternative models which could help support their validity.}

**Disclosure of assistance:** _Besides class materials, what other sources of assistance did you use while completing this lab? These can include input from classmates, relevant material identified through web searches (e.g., Stack Overflow), or assistance from ChatGPT or other AI tools. How did these sources support your own learning in completing this lab?_

{I got an error on question 2 regarding a sizing format that I had to run through ChatGPT. Specifically "Legend labels were too wide. The labels have been resized to 0.24, 0.24, 0.24, 0.24, 0.24. Increase legend.width (argument of tm_layout) to make the legend wider and therefore the labels larger."}

**Lab reflection:** _How do you feel about the work you did on this lab? Was it easy, moderate, or hard? What were the biggest things you learned by completing it?_

{This lab was fun. I enjoyed learning about the application of an NDVI in R. Although, I did have some very odd errors I had to troublshoot through.}